---
title: "Lake Washington metagenome and metatranscriptome meta-analysis"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

I have downloaded metagenomes and metatranscriptomes from the JGI IMG database for the following studies:

-   "Freshwater sediment methanotrophic microbial communities from Lake Washington under simulated oxygen tension"

-   "Synthetic bacterial communities of pure cultures isolated from sediment of Lake Washington, USA"

I want to take a look at these files since I don't really know what's in the giant .tar file I have.

To work with .tar files in R, you can use the tar function to list the contents of the tar archive and the untar function to extract specific files without needing to extract the entire archive. This approach allows you to "peek" into the files contained within the .tar file. Here’s how you can do it:

Listing Contents of a .tar File
To list the contents of a .tar file, you can use the following. This command prints the names of the files contained in the tar archive to the console.
```{r}
tarfile <- "LW_QS_img_data.tar"
tar("tf", tarfile)
```

The error you're encountering with the tar function in R, indicating that the file size is limited to 8GB, suggests you're dealing with a limitation of the R implementation or possibly an underlying system limitation related to handling large files.

To work around this issue, you might consider using command-line tools directly from R using the system or system2 function, which allows you to execute external commands. Many systems' native tar command does not have the same limitation and can handle larger files. Here's how you can do it:

1. Listing Contents of a .tar File
To list the contents of the .tar file using the system's native tar command:
```{r}
tarfile <- "LW_QS_img_data.tar"
cmd <- paste("tar -tf", shQuote(tarfile))
system(cmd)
```

Looks like everything is listed as the IMG genome ID, so as an example: 2634166257.tar.gz


Extracting Specific Files to Peek into Their Contents
If you want to look at the contents of specific files within the .tar archive without extracting everything, you can use the untar function with the list and exdir parameters. However, to directly read the contents into R (for example, to read the first few lines of a file to understand its structure), you'll need a slightly more involved approach.

2. Extracting Files
If you want to extract one or more specific files to peek at their contents without fully extracting the .tar archive, you can also do this with system:

```{r}
# Specify the file(s) you want to extract
MGX_37_LOW7 <- "3300004164.tar.gz"
# Specify where to extract the files
extractionPath <- "genomes"

# Create the command string
cmd <- paste("tar -xf", shQuote(tarfile), "-C", shQuote(extractionPath), shQuote(MGX_37_LOW7))

# Execute the command
system(cmd)
```

List the files that are contained within 3300004164.tar.gz
```{r}
# List contents of the .tar.gz file
tarfile <- "genomes/3300004164.tar.gz"
cmd <- paste("tar -tzf", shQuote(tarfile))
system(cmd)
```
List of files within one metagenome.tar.gz file: 
  3300004164/3300004164.a.cog.txt
  3300004164/3300004164.a.depth.txt
  3300004164/3300004164.a.ec.txt
  3300004164/3300004164.a.faa
  3300004164/3300004164.a.fna
  3300004164/3300004164.a.gene_product.txt
  3300004164/3300004164.a.gff
  3300004164/3300004164.a.ko.txt
  3300004164/3300004164.a.map.txt
  3300004164/3300004164.a.pfam.txt
  3300004164/3300004164.a.phylodist.txt
  3300004164/3300004164.config
  3300004164/3300004164.crispr.txt
  3300004164/3300004164.u.cog.txt
  3300004164/3300004164.u.ec.txt
  3300004164/3300004164.u.faa
  3300004164/3300004164.u.fna
  3300004164/3300004164.u.gene_product.txt
  3300004164/3300004164.u.gff
  3300004164/3300004164.u.ko.txt
  3300004164/3300004164.u.map.txt
  3300004164/3300004164.u.pfam.txt
  3300004164/3300004164.u.phylodist.txt
  3300004164/README.txt

Look at specific files by extracting them into a folder
```{r}
# Extract a specific file from the .tar.gz file
tarfile <- "genomes/3300004164.tar.gz"
destDir <- "C:/Users/Delaney/OneDrive/unknown/Documents/University of Utah/Puri Lab/4. Lake Washington/LW_QS_R/Lake_Washington_QS"  # Make sure to use forward slashes or double backslashes
fileName <- "3300004164/README.txt"

cmd <- paste("tar -xzf", shQuote(tarfile), "-C", shQuote(destDir), shQuote(fileName))
system(cmd)
```

Looks like ".u." means "unassembled" and ".a." means "assembled"

Extract all the files that are assembled, as well as the .config and README, etc. 
```{r}
# Define the path to the .tar.gz file
tarGzFile <- "genomes/3300004164.tar.gz"

# List all files in the .tar.gz archive
cmdList <- sprintf("tar -tzf %s", shQuote(tarGzFile))
allFiles <- system(cmdList, intern = TRUE)

# Filter out files containing '.u.'
filesToExtract <- allFiles[!grepl("\\.u\\.", allFiles)]

# Extract these files
# Define the extraction directory
extractionDir <- "C:/Users/Delaney/OneDrive/unknown/Documents/University of Utah/Puri Lab/4. Lake Washington/LW_QS_R/Lake_Washington_QS"
dir.create(extractionDir, recursive = TRUE, showWarnings = FALSE)

# Iterate over files to extract and extract each
for (file in filesToExtract) {
  cmdExtract <- sprintf("tar -xzf %s -C %s %s", shQuote(tarGzFile), shQuote(extractionDir), file)
  system(cmdExtract)
}
```

